Dynamical modeling of tidal streams

### 2014-01-07 tex/ms.tex
streams: re-arrange, looks okay

\begin{abstract} 
  I present a new framework for modeling the dynamics of cold tidal
  streams. The framework consists of simple models for the initial
  action--angle distribution of tidal debris, which can be
  straightforwardly evolved forward in time. Taking advantage of the
  essentially one-dimensional nature of cold tidal streams, the
  transformation to position--velocity coordinates can be linearized
  and interpolated near a small number of points along the progenitor
  orbit, thus allowing for efficient computations of a stream's
  properties in observable quantities. I illustrate how to calculate
  the stream's average location (its ``track'') in different
  coordinate systems, how to quickly estimate the dispersion around
  its track, and how to draw mock stream data. As a generative model,
  this framework allows one to compute the full probability
  distribution function and marginalize over or condition it on
  certain phase--space dimensions as well as convolve it with
  observational uncertainties. This will be instrumental in proper
  data analysis of stream data. In addition to providing a
  computationally-efficient practical tool for modeling the dynamics
  of tidal streams, the action--angle nature of the framework helps
  elucidate in exactly what manner streams do not follow single
  orbits, how the observed width of the stream relates to the velocity
  dispersion or mass of the progenitor, and how the progenitors of
  ``orphan'' streams could be located.

@@

In this paper, I present a new method for modeling cold tidal streams
that fundamentally lives in frequency--angle space, because the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle offset
distributions are independent of stripping time. While this model
oversimplifies the real stripping process, I argue that at least for
constraining the gravitational potential with stream data, this simple
model is likely unbiased. For detailed mock stream data, more
elaborate models of the stripping process as a function of time might
be necessary and I discuss how these could easily be incorporated into
the proposed framework.

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa}. This transformation is linearized at a small
number of points along the progenitor orbit that span the length of
the stream. For cold streams, the separation between the stream track
and the progenitor orbit as well as the internal differences within
the stream are small, such that this linear approximation is accurate
enough. The linear approximation allows for fast evaluation of the
average stream location (the stream ``track'') in position--velocity
coordinates and efficient mock data generation. For a uniform
distribution of stripping times, an observed stream member's
phase--space probability distribution function (PDF) can be
analytically marginalized over stripping time, such that using the
linear approximation any evaluation---including marginalization and
uncertainty convolution---of the PDF is extremely rapid. This clears
the way for proper probabilistic inference of the Milky Way halo's
gravitational potential and progenitor properties using stream data
for individual stars.

@@

\begin{figure}[t!]
  \includegraphics[width=0.48\textwidth,clip=]{gd1_evol_aadohist.eps}
  \caption{Distribution of the magnitude of frequency offsets between
    stream particles and the progenitor along the direction of the
    mean frequency offset. The dashed line is a Gaussian fit to this
    distribution, while the solid line is a fit of our preferred form
    for this distribution, given in \equationname~(\ref{eq:do}) (the
    dashed line is almost entirely covered by the solid line). Both
    forms provide adequate fits to the
    distribution. }\label{fig:gd1_dohist}
%python plot_stream.py ../tex/gd1_evol_aadohist.ps
\end{figure}

@@

Because we expect the typical $\Delta\veco$ to scale with the velocity
dispersion of the progenitor, I conjecture that a constant
$\mu_\Omega$ can be used for modeling cold tidal streams of any
(small-ish) velocity dispersion. Further modeling of simulated
streams, however, is necessary to check this and to find the optimal
value of $\mu_\Omega$. In particular, if the progenitor has internal
rotation the action distribution of the debris and $\mu_\Omega$ may be
significantly different. Fixing $\mu_\Omega$ leaves a single free
parameter \sigv\ for modeling the frequency offset distribution of a
tidal stream. In modeling to observed data, this parameter needs to be
fit and we expect it to be proportional to the velocity dispersion of
the progenitor, although this proportionality needs to be checked more
carefully (but see Figure 3 in \citealt{Sanders13a}, which shows that
the size of the frequency distribution scales as mass$^{1/3}$, or
approximately as the velocity dispersion). Similarly, the initial
angle distribution, modeled here as an isotropic Gaussian, has a
characteristic spread $\sigma_{\theta}$ which I set here to $0.003$
($=\sigv/[122\kms]$), and we likewise expect this spread to scale with
the velocity dispersion of the progenitor.

@@

@@

@@

\begin{figure*}[t!]
\begin{center}
  \includegraphics[width=0.36\textwidth,clip=]{gd1_evol_aaaparopar.ps}
  \includegraphics[width=0.36\textwidth,clip=]{gd1_evol_aaaparoperp.ps}\\
  \includegraphics[width=0.36\textwidth,clip=]{gd1_evol_aaaparaperp.ps}
  \includegraphics[width=0.36\textwidth,clip=]{gd1_evol_aaapartime.ps}
  \caption{Stream properties as a function of angle along the
    stream. The top, left panel shows the magnitude of frequency
    offsets along the stream. The gray line shows the mean predicted
    frequency offset calculated using the framework of
    \sectionname\sectionname~\ref{sec:modeloa} and \ref{sec:track},
    while the grayscale bands show the $68\,\%$ and $95\,\%$ limits of
    the predicted distribution. The dashed line shows the angle $\apar
    = \Delta \Omega^m\,t_d$, angles beyond which can only be reached
    by having a larger $\Delta \opar$ than average. The top, right
    panel shows the magnitude of frequency offsets perpendicular to
    those in the left panel. The predicted distribution is constant as
    a function of angle and therefore not shown. The bottom, left
    panel shows the angle offset perpendicular to the stream and the
    grayscale bands show $68\,\%$ and $95\,\%$ limits of the predicted
    distribution of angle offsets; the white dashed line shows the
    $68\,\%$ limit calculated using a simplified method. The bottom,
    right panel shows the stripping time. The gray line and grayscale
    bands again show the mean and $68\,\%$ and $95\,\%$ confidence
    limits of the predicted distribution. While the ``burstiness''
    related to the energetic stripping at pericenter that is apparent
    in the simulated data's distributions is (by design) absent in the
    model, the overall scale and the mean trends with angle are well
    represented by the model.}\label{fig:gd1_apar}
%python plot_stream.py ../tex/gd1_evol_aaaparopar.ps
%python plot_stream.py ../tex/gd1_evol_aaaparoperp.ps
%python plot_stream.py ../tex/gd1_evol_aaaparaperp.ps
%python plot_stream.py ../tex/gd1_evol_aaapartime.ps
\end{center}
\end{figure*}

@@

\begin{figure*}
\begin{center}
  \includegraphics[width=0.8\textwidth,clip=]{track_aa.eps}\\
  \includegraphics[width=0.8\textwidth,clip=]{track_xz.eps}\\
  \includegraphics[width=0.8\textwidth,clip=]{track_ld.eps}
  \caption{Illustration of the determination of the average stream
    track described in \sectionname~\ref{sec:track}. The small black
    dots in each panel are the points along the progenitor track at
    which the Jacobian $\partial (\veco,\veca) / \partial
    (\vecx,\vecv)$ is calculated. The large black dots are the
    corresponding points on the stream where the track is
    calculated. The dashed line in each panel is the progenitor's
    orbit, while the solid line and dash-dotted lines are the mean
    track and $2\sigma$ dispersion.}\label{fig:track}
%python illustrate_track.py ../tex/track_aa.ps ../tex/track_xz.ps ../tex/track_ld.ps
\end{center}
\end{figure*}

@@

\begin{figure*}[th!!]
\begin{center}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_lb.eps}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_ld.eps}
  \caption{Simulated stream after $5.011\Gyr$ as a function of
    Galactic longitude. The top panel shows the stream in Galactic
    latitude and the bottom panel shows it in the distance from the
    Sun. The progenitor is shown as the large gray dot and its orbit
    as the dashed lines. The predicted track of the stream is shown as
    the solid line. Insets show various zoomed-in parts of the stream
    to show the stream, progenitor orbit, and predicted stream track
    up close.}\label{fig:gd1_lbd}
%python plot_stream.py ../tex/gd1_evol_lb.ps
%python plot_stream.py ../tex/gd1_evol_ld.ps
\end{center}
\end{figure*}

Having estimated the track of the stream and the dispersion around
this in frequency--angle space, we can propagate the track to
position--velocity coordinates by inverting the transformation
$(\vecx,\vecv) \rightarrow (\veco,\veca)$. I do this by linearizing
the transformation along the orbit of the progenitor in the vicinity
of the stream and inverting this linearized transformation. That this
works relies on the stream being cold for a few different
reasons. First, the track of a cold stream, while in general offset
from the orbit of the progenitor, does not stray too far from it, such
that the transformation to frequency--angle coordinates can be
linearized over the range of the stream--orbit offset. Second, cold
streams are essentially one-dimensional, that is, they only spread
significantly over a single direction. This means that non-linearity
in the $(\vecx,\vecv) \rightarrow (\veco,\veca)$ transformation only
affects a single direction, which roughly follows the orbit of the
progenitor. Therefore, we can linearize the $(\vecx,\vecv) \rightarrow
(\veco,\veca)$ transformation along a one-dimensional grid of points
on the progenitor's orbit, rather than on a full six-dimensional
grid. Third, the coldness of the stream means that any significant
mass of the full stream PDF is close enough to the stream track that
all frequency--angle calculations can be performed using the linear
approximation along the progenitor orbit.

>>>>

\begin{abstract} 
  I present a new framework for modeling the dynamics of tidal
  streams. The framework consists of simple models for the initial
  action--angle distribution of tidal debris, which can be
  straightforwardly evolved forward in time. Taking advantage of the
  essentially one-dimensional nature of tidal streams, the
  transformation to position--velocity coordinates can be linearized
  and interpolated near a small number of points along the progenitor
  orbit, thus allowing for efficient computations of a stream's
  properties in observable quantities. I illustrate how to calculate
  the stream's average location (its ``track'') in different
  coordinate systems, how to quickly estimate the dispersion around
  its track, and how to draw mock stream data. As a generative model,
  this framework allows one to compute the full probability
  distribution function and marginalize over or condition it on
  certain phase--space dimensions as well as convolve it with
  observational uncertainties. This will be instrumental in proper
  data analysis of stream data. In addition to providing a
  computationally-efficient practical tool for modeling the dynamics
  of tidal streams, the action--angle nature of the framework helps
  elucidate in exactly what manner streams do not follow single
  orbits, how the observed width of the stream relates to the velocity
  dispersion or mass of the progenitor, and how the progenitors of
  ``orphan'' streams could be located.

@@

In this paper, I present a new method for modeling tidal streams that
fundamentally lives in frequency--angle space, because the stream
distribution function is essentially one-dimensional in this
space. While many of the results of this paper are more generally
valid, the fiducial stream model in this paper consists of a
three-dimensional (close to) Gaussian distribution of frequencies in
the stream, a three-dimensional Gaussian distribution of initial angle
offsets between stream members and the progenitor, and a uniform
distribution of stripping times; the frequency and angle offset
distributions are independent of stripping time. While this model
oversimplifies the real stripping process, I argue that at least for
constraining the gravitational potential with stream data, this simple
model is likely unbiased. For detailed mock stream data, more
elaborate models of the stripping process as a function of time might
be necessary and I discuss how these could easily be incorporated into
the proposed framework.

To evaluate the model in position--velocity coordinates, the
transformation to and from action--angle coordinates is computed using
a novel method for calculating action--angle coordinates presented in
Appendix~\ref{sec:aa}. This transformation is linearized at a small
number of points along the progenitor orbit that span the length of
the stream. For streams originating from progenitors with masses
$\lesssim 10^8\msun$, the separation between the stream track and the
progenitor orbit as well as the internal differences within the stream
are small, such that this linear approximation is accurate enough. The
linear approximation allows for fast evaluation of the average stream
location (the stream ``track'') in position--velocity coordinates and
efficient mock data generation. For a uniform distribution of
stripping times, an observed stream member's phase--space probability
distribution function (PDF) can be analytically marginalized over
stripping time, such that using the linear approximation any
evaluation---including marginalization and uncertainty
convolution---of the PDF is extremely rapid. This clears the way for
proper probabilistic inference of the Milky Way halo's gravitational
potential and progenitor properties using stream data for individual
stars.

@@

\begin{figure}[t!]
  \includegraphics[width=0.44\textwidth,clip=]{gd1_evol_aadohist.eps}
  \caption{Distribution of the magnitude of frequency offsets between
    stream particles and the progenitor along the direction of the
    mean frequency offset. The dashed line is a Gaussian fit to this
    distribution, while the solid line is a fit of our preferred form
    for this distribution, given in \equationname~(\ref{eq:do}) (the
    dashed line is almost entirely covered by the solid line). Both
    forms provide adequate fits to the
    distribution. }\label{fig:gd1_dohist}
%python plot_stream.py ../tex/gd1_evol_aadohist.ps
\end{figure}

@@

Because we expect the typical $\Delta\veco$ to scale with the velocity
dispersion of the progenitor, I conjecture that a constant
$\mu_\Omega$ can be used for modeling tidal streams of any (small-ish)
velocity dispersion. Further modeling of simulated streams, however,
is necessary to check this and to find the optimal value of
$\mu_\Omega$. In particular, if the progenitor has internal rotation
the action distribution of the debris and $\mu_\Omega$ may be
significantly different. Fixing $\mu_\Omega$ leaves a single free
parameter \sigv\ for modeling the frequency offset distribution of a
tidal stream. In modeling to observed data, this parameter needs to be
fit and we expect it to be proportional to the velocity dispersion of
the progenitor, although this proportionality needs to be checked more
carefully (but see Figure 3 in \citealt{Sanders13a}, which shows that
the size of the frequency distribution scales as mass$^{1/3}$, or
approximately as the velocity dispersion). Similarly, the initial
angle distribution, modeled here as an isotropic Gaussian, has a
characteristic spread $\sigma_{\theta}$ which I set here to $0.003$
($=\sigv/[122\kms]$), and we likewise expect this spread to scale with
the velocity dispersion of the progenitor.

@@

\begin{figure*}[t!]
\begin{center}
  \includegraphics[width=0.34\textwidth,clip=]{gd1_evol_aaaparopar.eps}
  \includegraphics[width=0.34\textwidth,clip=]{gd1_evol_aaaparoperp.eps}\\
  \includegraphics[width=0.34\textwidth,clip=]{gd1_evol_aaaparaperp.eps}
  \includegraphics[width=0.34\textwidth,clip=]{gd1_evol_aaapartime.eps}
  \caption{Stream properties as a function of angle along the
    stream. The top, left panel shows the magnitude of frequency
    offsets along the stream. The gray line shows the mean predicted
    frequency offset calculated using the framework of
    \sectionname\sectionname~\ref{sec:modeloa} and \ref{sec:track},
    while the grayscale bands show the $68\,\%$ and $95\,\%$ limits of
    the predicted distribution. The dashed line shows the angle $\apar
    = \Delta \Omega^m\,t_d$, angles beyond which can only be reached
    by having a larger $\Delta \opar$ than average. The top, right
    panel shows the magnitude of frequency offsets perpendicular to
    those in the left panel. The predicted distribution is constant as
    a function of angle and therefore not shown. The bottom, left
    panel shows the angle offset perpendicular to the stream and the
    grayscale bands show $68\,\%$ and $95\,\%$ limits of the predicted
    distribution of angle offsets; the white dashed line shows the
    $68\,\%$ limit calculated using a simplified method. The bottom,
    right panel shows the stripping time. The gray line and grayscale
    bands again show the mean and $68\,\%$ and $95\,\%$ confidence
    limits of the predicted distribution. While the ``burstiness''
    related to the energetic stripping at pericenter that is apparent
    in the simulated data's distributions is (by design) absent in the
    model, the overall scale and the mean trends with angle are well
    represented by the model.}\label{fig:gd1_apar}
%python plot_stream.py ../tex/gd1_evol_aaaparopar.ps
%python plot_stream.py ../tex/gd1_evol_aaaparoperp.ps
%python plot_stream.py ../tex/gd1_evol_aaaparaperp.ps
%python plot_stream.py ../tex/gd1_evol_aaapartime.ps
\end{center}
\end{figure*}

@@

\begin{figure*}[t!p]
\begin{center}
  \includegraphics[width=0.8\textwidth,clip=]{track_aa.eps}\\
  \includegraphics[width=0.8\textwidth,clip=]{track_xz.eps}\\
  \includegraphics[width=0.8\textwidth,clip=]{track_ld.eps}
  \caption{Illustration of the determination of the average stream
    track described in \sectionname~\ref{sec:track}. The small black
    dots in each panel are the points along the progenitor track at
    which the Jacobian $\partial (\veco,\veca) / \partial
    (\vecx,\vecv)$ is calculated. The large black dots are the
    corresponding points on the stream where the track is
    calculated. The dashed line in each panel is the progenitor's
    orbit, while the solid line and dash-dotted lines are the mean
    track and $2\sigma$ dispersion.}\label{fig:track}
%python illustrate_track.py ../tex/track_aa.ps ../tex/track_xz.ps ../tex/track_ld.ps
\end{center}
\end{figure*}

@@

@@

\begin{figure*}[t!]
\begin{center}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_lb.eps}
  \includegraphics[width=0.8\textwidth,clip=]{gd1_evol_ld.eps}
  \caption{Simulated stream after $5.011\Gyr$ as a function of
    Galactic longitude. The top panel shows the stream in Galactic
    latitude and the bottom panel shows it in the distance from the
    Sun. The progenitor is shown as the large gray dot and its orbit
    as the dashed lines. The predicted track of the stream is shown as
    the solid line. Insets show various zoomed-in parts of the stream
    to show the stream, progenitor orbit, and predicted stream track
    up close.}\label{fig:gd1_lbd}
%python plot_stream.py ../tex/gd1_evol_lb.ps
%python plot_stream.py ../tex/gd1_evol_ld.ps
\end{center}
\end{figure*}

@@

Having estimated the track of the stream and the dispersion around
this in frequency--angle space, we can propagate the track to
position--velocity coordinates by inverting the transformation
$(\vecx,\vecv) \rightarrow (\veco,\veca)$. I do this by linearizing
the transformation along the orbit of the progenitor in the vicinity
of the stream and inverting this linearized transformation. That this
works relies on the stream being relatively cold (in that $\sigv/V_c
\ll 1$) for a few different reasons. First, the track of a cold
stream, while in general offset from the orbit of the progenitor, does
not stray too far from it, such that the transformation to
frequency--angle coordinates can be linearized over the range of the
stream--orbit offset. Second, cold streams are essentially
one-dimensional, that is, they only spread significantly over a single
direction. This means that non-linearity in the $(\vecx,\vecv)
\rightarrow (\veco,\veca)$ transformation only affects a single
direction, which roughly follows the orbit of the
progenitor. Therefore, we can linearize the $(\vecx,\vecv) \rightarrow
(\veco,\veca)$ transformation along a one-dimensional grid of points
on the progenitor's orbit, rather than on a full six-dimensional
grid. Third, the coldness of the stream means that any significant
mass of the full stream PDF is close enough to the stream track that
all frequency--angle calculations can be performed using the linear
approximation along the progenitor orbit.
